Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-02-05 papers/LOPe/intro.tex

For the purpose of discussion, we define a degree of difficulty term called the joint Parallel
Complexity Product (PCP),

\begin{verbatim}
    PCP = code_complexity X compiler_complexity
\end{verbatim}
and suggest that PCP is roughly a constant over time, say the level of complexity of writing an MPI
program today, PCP\_MPI.

Of course one would hope that over time code complexity goes down as better languages allow
compilers to take on a greater share of the parallel complexity burden.  This is somewhat true of
Unified Parallel C (UPC) and Coarray Fortran.  With the PGAS extensions, C and Fortran compilers
are now aware of parallelism and now generate message passing code that previously had been handled
by the MPI library.  In some instances the compiler is able to perform optimizations that it is not
able to do with a library-based scheme like MPI \cite{preissl:2011:SC}.

@@

However, in large part these languages are largely syntatic sugar for message passing and do not
provide a substatial decrease in code complexity when application performance is a
goal\cite{hasert:2011:EuroMPI}.  While skilled programmers in the HPC community have become
accustomed to the level of complexity of an MPI program,
%%and yet are welcome to considering the simplifications of UPC and Coarray Fortran.
the problem for programmers is that hardware is changing in ways that increase the level of on-chip
parallelism.  Indeed, the current generation of machines could be the last to provide homogeneous
multi-core parallelism\cite{JORSjors.aw}.  For ultimate performance, programmers must envision
coding for heterogeneous architectures 
%%hyper-parallelism
with huge new levels of on-node parallelism, at the same time they account for off-node
parallelism.  Thus PCP has suddenly increased with PCP $=$ PCP\_MPI $+$ PCP\_Multi\_Core $>>$
PCP\_MPI.  Since languages have not evolved that allow the compiler to take up this increase, the
complexity for a programmer has necessarily dramatically increased.  Unfortunately, considering the
complexity of large multiphysics code frameworks, it could take several person years to retarget an
application for \emph{each} new heterogeneous platform\cite{DBLP:journals/corr/DubeyS13}.

@@

In this paper we examine a language-based paradigm that allows the compiler to take on a larger
portion of the PCP burden.  Anyone programming in OpenCL or CUDA is aware that explicit loop
structures over array elements in a serial program are removed and replaced by a kernel program
that is run over all of the elements of the input arrays.  We propose Locally Orientated
Programming extensions (LOPe) to the Fortran and potentially C languages that formally adopt this
programming model.  The LOPe programming model explicitly separates numerical complexity from
parallel complexity --- leaving much of the staging and assembling of data to the compiler --- and
is applicable to the solution of stencil-based algorithms that provide data access patterns that
are regular and spatially local.

>>>>

Of course one would hope that over time code complexity goes down as
better languages allow compilers to take on a greater share of the
parallel complexity burden.  Multithreaded code written using OpenMP
can be significantly simpler than the corresponding code in which the
programmer explicitly manages a pool of threads themselves.  This is
somewhat true for distributed memory parallelism using Unified
Parallel C (UPC) and Coarray Fortran.  With the PGAS extensions, C and
Fortran compilers are now aware of parallelism and now generate
message passing code that previously had been handled by the MPI
library.  In some instances the compiler is able to perform
optimizations that it is not able to do with a library-based scheme
like MPI \cite{preissl:2011:SC}.

@@

However, in large part these languages are largely syntatic sugar for
message passing and do not provide a substantial decrease in code
complexity when application performance is a
goal\cite{hasert:2011:EuroMPI}.  While skilled programmers in the HPC
community have become accustomed to the level of complexity of an MPI
program, the problem for programmers is that hardware is changing in
ways that increase the level of on-chip parallelism.  Indeed, the
current generation of machines could be the last to provide
homogeneous multi-core parallelism\cite{JORSjors.aw}.  For ultimate
performance, programmers must envision coding for heterogeneous
architectures with huge new levels of on-node parallelism, at the same
time they account for off-node parallelism.
Since languages have not evolved that allow the compiler to take up
this increase, the complexity for a programmer has necessarily
dramatically increased.  Unfortunately, considering the complexity of
large multiphysics code frameworks, it could take several person years
to retarget an application for \emph{each} new heterogeneous
platform\cite{DBLP:journals/corr/DubeyS13}.

@@

In this paper we examine a language-based paradigm that allows the
compiler to take on a larger portion of the code complexity burden.
Anyone programming in OpenCL or CUDA is aware that explicit loop
structures over array elements in a serial program are removed and
replaced by a kernel program that is run over all of the elements of
the input arrays.  We propose \emph{Locally Orientated Programming
extensions} (LOPe) to the Fortran and potentially C languages that
formally adopt this programming model.  The LOPe programming model
explicitly separates numerical complexity from parallel complexity ---
leaving much of the staging and assembling of data to the compiler ---
and is applicable to the solution of stencil-based algorithms that
provide data access patterns that are regular and spatially local.
